---
title: "TMA4315: Compulsory exercise 2" 
subtitle: "Aurora Hofman, Camilla Karlsen, Catharina Lilleengen" 
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: html_document
---


# Problem 1

```{r setup, include=TRUE}
  filepath <- "https://www.math.ntnu.no/emner/TMA4315/2018h/mountains"
  mount <- read.table(file = filepath, header = TRUE, col.names = c("height", 
      "prominence", "fail", "success"))

  height <- mount$height
  prominence <- mount$prominence
  success <- mount$success
  fail <- mount$fail
```
##a) 
```{r, include=TRUE}
  mod <- glm(cbind(success,fail)~height+prominence, family=binomial(link = "logit"))
  summary(mod)
```
Here we want to fit a binary regression 

Catharina skriver inn denne senere

##b) 
By used of the observed deviance, we can estimate the overdispersion parameter $\phi$. We have, 
$$
  \hat{\phi} = \frac{D}{G-p}, 
$$
where $D$ is the deviance, $G$ is the number of groups and $p$ is the number of estimated regression coefficients. 
```{r, include=TRUE}
  deviance <- summary(mod)$deviance
  G <- nrow(mount)      #number of groups
  p <- ncol(mount)-1    #number of estimated regression coefficients
  phi <- deviance/(G-p)
```
Since $phi$ is larger than 1 we have overdispersion. Reasons for overdispersion can be unobserved heterogeneity and positive correlation between independent binary observation of the response variable. In this case, the data will be correlated because each individual most likely will belong to a cluster since reaching a mountain often happens in groups with a guide. Because we have overdispersion of we want to refit the model using a quasi-likelihood model. 

```{r, include=TRUE}
  mod2 <- glm(cbind(success,fail)~height+prominence, family=quasibinomial(link = "logit"))
  summary(mod2)
```

##c) 
```{r, include=TRUE}
  library(ISwR)
  #install.packages("ISwR")

  #Model selection with QAIC  
  #Full model
  likelihood = logLik(mod)
  QAIC_mod = -2*likelihood/phi+2*(p+1)
  likelihood
  QAIC_mod
  
  #Model with only height as covariate
  mod10 <- glm(cbind(success,fail)~height, family=binomial(link = "logit"))
  deviance10 <- summary(mod10)$deviance
  phi10 <- deviance10/(G-p-1)
  likelihood10 = logLik(mod10)
  QAIC_mod10 = -2*likelihood10/phi10+2*(p)
  likelihood10
  QAIC_mod10
  
  #Model with only prominence as covariate
  mod01 <- glm(cbind(success,fail)~prominence , family=binomial(link = "logit"))
  deviance01 <- summary(mod01)$deviance
  phi01 <- deviance01/(G-p-1)
  likelihood01 = logLik(mod01)
  QAIC_mod01 = -2*likelihood01/phi01+2*(p)
  likelihood01
  QAIC_mod01
  
  #Model with only intercept
  mod00 <- glm(cbind(success,fail)~1 , family=binomial(link = "logit"))
  deviance00 <- summary(mod00)$deviance
  phi00 <- deviance00/(G-p-2)
  likelihood00 = logLik(mod00)
  QAIC_mod00 = -2*likelihood00/phi00+2*(p-1)
  likelihood00
  QAIC_mod00
  
  #SKAL VI OGSÅ HA MED AIC FOR HVER MODEL HER??
```
By looking at the computed QAICs we see that the model with only intercept has the lowest value. Hence, we choose this model without any covariates as our model. 

##d)
```{r, include=TRUE}
drop1(mod, test = "LRT")  #Sett inn modellen vi valgte over


#Wald test
beta <- coef(mod)
d <- c(0,0,0)
C <- rbind(c(1,0,0),
           c(0,1,0),
           c(0,0,1))
#Hvordan skal C se ut???
wald <- t(C %*% beta - d) %*% solve(C %*% vcov(mod) %*% t(C)) %*% (C %*% beta - d)
wald
pchisq(wald, df=2, lower.tail=FALSE)

  
```
By use of likelihood ratio test we see that each term is significant in the model. By use of the Wald test 

##e) 
```{r, include=TRUE}
#residuals(,type = "deviance")
  
```


##f) 
```{r, include=TRUE}

  
```

# Problem 2

```{r, include=TRUE}
 long <- read.csv("https://www.math.ntnu.no/emner/TMA4315/2019h/eliteserie.csv")
```

##a) 

##b) 

##c) 

##d) 

##e) 









